---
title: "Basic workflow with talkr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(talkr)
```

## Data

We will be using the IFADV corpus as example data for the workflow of `talkr`.
A prepared dataset can be downloaded by installing the `ifadv` package:

```{r install data package}
# install.packages("devtools")
devtools::install_github("elpaco-escience/ifadv")
```

We will load the ifadv dataset in the object `data`, as follows:

```{r}
data <- ifadv::ifadv
```

## Summaries

The `report_summaries` function provides a summary of the data, including the
total number of utterances, the total duration of the conversation, the number
of speakers, and the number of sources.

```{r}
report_summaries(data)
```

## Visual quality checks

The `plot_quality` function provides a visual check of the quality of the data.
