---
title: "Basic workflow with talkr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(talkr)
```

## Data

We will be using the IFADV corpus as example data for the workflow of `talkr`.
A prepared dataset can be downloaded by installing the `ifadv` package:

```{r install data package}
# install.packages("devtools")
devtools::install_github("elpaco-escience/ifadv")
```

We will initialize the talkr dataset using the ifadv data, as follows:

```{r}
data <- init(ifadv::ifadv)
```

For the `talkr` workflow, the columns named `source`, `begin`, `end`, `participant` and `utterance` are essential.
The `init()` function can be used to rename columns if necessary.
For example, if the column `participant` is named `speaker`, we can rename it as follows:

```r
talkr_data <- init(data,
             participant = "speaker")
```

## Summaries

The `report_summaries` function provides a summary of the data, including the
total number of utterances, the total duration of the conversation, the number
of speakers, and the number of sources.

```{r}
report_stats(data)
```

## Visual quality checks

The `plot_quality` function provides a visual check of the quality of the data, by visualizing the distribution of turn duration, and transition timing.

Transition timing is similar to FTO, but calculated without additional quality checks: transitions are identified when the participant changes from one turn to the next. The transition time is then calculated as the difference between the beginning of the turn of the new participant, and the end of the turn of the previous one.

By default, `plot_quality()` will plot the quality of the entire dataset:

```{r}
plot_quality(data)

```

Quality plots can also be run for a specific source:

```{r}
plot_quality(data, source = "/dutch2/DVA9M")

```

## Plot conversations

Individual conversations can be plotted quickly using `plot_turns_tokens()`.
The default setting is to plot the first 60 seconds of the first source in the data,
overlaying the 10 most frequent tokens.

```{r}
plot_turns_tokens(data)
```

We can set other defaults; e.g. a specific source, a different time window, and a different number of tokens:

```{r}
plot_turns_tokens(data, source = "/dutch2/DVA9M",
                  begin = 120,
                  duration = 120,
                  maxrank = 20)
```

For more control over the plot, two specific geometries are available: `geom_turn` and `geom_token`.
In addition, there is a `talkr`-specific theme provided.

```{r}
p <- data |>
  dplyr::filter(source == "/dutch2/DVA9M") |>
  dplyr::filter(end < 60000) |>
  ggplot2::ggplot(aes(x = end, y = participant)) +
  geom_turn(aes(
    begin = begin,
    end = end)) +
  xlab("Time (ms)") +
  ylab("") +
  theme_turnPlot()

p
```

This plot can be overlayed with plotted occurrences of tokens.

To do so, we first need to calculate the token frequencies:

```{r}
tokens <- tokenize(data)

tokens
```

Token frequencies are calculated over the entire dataset. For source-specific data, it is recommended to filter
the source prior to tokenization:

```{r}
tokens <- data |>
  dplyr::filter(source == "/dutch2/DVA9M") |>
  tokenize()

tokens
```

Before we plot the tokens over the turns, we need to select the tokens we want to plot (e.g. the top 10 ranked), and the time window they occur in:

```{r}
tokenselection <- tokens |>
  dplyr::filter(relative_time < 60000) |>
  dplyr::filter(rank <= 10)
```

We can plot the tokens over the turns.

```{r}
p +
geom_token(data = tokenselection,
           aes(x = relative_time,
               y = participant,
               color = rank)) +
  viridis::scale_color_viridis(option = "plasma", direction = -1, begin = 0.2, end = 0.8)
```
